Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study

### 2024-03-18 manuscript.Rmd
18/3

---
title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
output: html_document
author: "Elin Sun Cao"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

This was a registry-based study using all patients between 2014 and 2022
from the Karolinska University Hospital in Solna who had been admitted
to the ICU and who were reviewed regarding the presence of OFI by the
mortality and morbidity conference. Bi-variable and multivariable
logistic regression was used to assess the following patient and process
factors and OFI among this group of patients: gender, age, if the
patient was intubated or not and how many days, respiratory rate,
systolic blood pressure, Glasgow Coma Scale (GCS), Injury Severity Score
(ISS), on call, time from arrival at the hospital until first computed
tomography (CT), ASA preinjury and survival after 30 days.

@@

@@

@@

@@

@@

RTS is a physiological scoring system scored from the data obtained in
the first set of data obtained on the patient and consists of Glasgow
Coma Scale (GCS), respiratory rate and systolic blood pressure. RTS is
heavily weighted towards GCS compared to respiratory rate and systolic
blood pressure. It has been demonstrated to show accuracy at predicting
death.

@@

We conducted a retrospective single-centre cohort study based on data
from the Karolinska University Hospital in Solna and from the Swedish
Trauma registry. We used R for all statistical analyses [@19]. For
patients where hospital values for systolic blood pressure, respiratory
rate or GCS score were missing, prehospital values were used instead. A
complete case analysis was conducted after handling missing values. We
present sample characteristics using descriptive statistics.

@@

The quantitative variables were factors chosen from the trauma registry,
based on the locally used audit filters, standard epidemiological
factors and factors registered in the Swedish Trauma Registry. The
categorical factors were sex, age, if the patient was intubated or not
and how many days, respiratory rate, systolic blood pressure, Glasgow
Coma Scale (GCS), Injury Severity Score (ISS), on call, time from
arrival at the hospital until first computed tomography (CT), ASA
preinjury and survival after 30 days. On duty was defined as before 8.00
a.m., after 5 p.m., or during a weekend - defined as Saturday or Sunday.

@@

#Detta blir upprepning då det redan står längre ner i resultatdelen?

@@

# Results
## Participants
```{r, include=FALSE}
source("functions/flowchart.R")
```
```{r}
for (i in 1:(length(vert) - 1)) {
  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
    print
}

@@

## Descriptive data
```{r include_scripts, echo=FALSE, include=FALSE}
source("functions/table1.R")
```
```{r}
table2
```

## Main results
```{r, include=FALSE}
source("functions/table2.R")
```
```{r}
table3_merge
```

14,022 patients were included in the trauma registry and trauma care
quality database between 2017 and 2023. 12,278 patients were excluded,
leaving a total of 1,742 patients eligible for the study. 11,343 were
excluded because they were not treated in the ICU. 3 patients were
excluded because they were under 15 years old. 6 patients were excluded
because they were dead on arrival, and 928 patients were excluded
because there was no data on the presence or absence OFI.

293 patients were excluded because there was missing data, leaving a
total of 1,449 patients included in the study.

Table 1 present sample characteristics. 1,096 (76%) patients were male and 353 (24%) patients were female. 846 (58%) patients had no intubation, 241 (17%) patients were intubated for 1-7 days and 113 (7,8%) patients were intubated for more than 7 days. The number of patients who arrived outside working hours — on Saturday or Sunday or after 5 pm or before 8 am were 1,042 (72%). 1,064 (73%) patients were equal to or less than 7 days in the ICU, and 385 (27%) patients were more than 7 days in the ICU. 1,080 (75%) patients had an ASA-score of 1-2 before arrival to the hospital, and 369 (25%) patients had an ASA-score of 3-7 before arrival to the hospital. 286 (20%) patients died within 30 days after hospitalization.

@@

RTS and ≤ 7 days in the ICU were significantly associated with increased odds of OFI in both the
unadjusted and adjusted analyses.

Sex, Age, Intubation, ISS, time to first CT, on duty and ASA preinjury were not significantly associated with OFI in the unadjusted or the adjusted analyses. Female sex, intubation > 7 days, > 7 days in the ICU, ASA preinjury 1-2 and survival 30 days after hospitalisation could not be analysed with logistic regression.

@@

We found that RTS and ≤ 7 days in the ICU were associated with higher odds of OFI. Previous
research has shown that ...

These findings could be because ...

Death within 30 days after hospitalisation was associated with OFI in the unadjusted analysis but not in the
adjusted analysis.

1.  add sources from previous studies

2.  studies to back up the results from the analyses

@@

There were several limitations in our study. The selection for review
for the morbidity and mortality review relied mostly on the local audit
filters used at the Karolinska Institute Hospital in Solna. This leads
to a risk of misplacement where some patients with OFI were not selected
by for the review based on the audit filters although there was an OFI.

Another limtation is that some relevant factors, such as ----, were not
part of the Swedish Trauma registry and therefore not adjusted for in
the analysis.

The study was a single-center study and illustrate the situation in
Stockholm. More study is needed on the area of OFI in patients admitted
to the ICU. A suggestion would be to conduct a multi-center study on
this group of patients to further investigate areas for improvement and
to further develop trauma quality programs.

Future studies could also focus on ---- which was seen to be associated
with OFI in these patients.

<!--Lägg till 2 st förslag på explicita fortsättningsstudier! -->

@@

The study is conducted on 143 outcomes of OFI. There were few outcomes because
we included only those who were admitted to the ICU, \> 14 years old,
not dead on arrival and had a registration for the presence or absence
of OFI. The external validity is therefore compromised due to the
limited size of the Swedish database. However, ...

@@

---- among patients admitted to the ICU have higher odds of OFI. These
factors can be seen as new targets for further reseach to further
supplement trauma quality improvement programs.

<!-- Give a short conclusion, corresponding to the aim -->

>>>>

---
title: "Opportunities for improvement in adult trauma patients admitted to the intensive care unit: A registry-based study"
output: "html_document"
author: "Elin Sun Cao"
bibliography: bibliography.bib
csl: vancouver.csl
editor_options: 
  markdown: 
    wrap: 72
---

@@

This was a registry-based study using all patients between 2014 and 2022
from the Karolinska University Hospital in Solna who had been admitted
to the ICU and who were reviewed regarding the presence of OFI by the
mortality and morbidity conference. Bi-variable and multivariable
logistic regression was used to assess the following patient and process
factors and OFI among this group of patients: gender, age, if the
patient was intubated or not and how many days, Revised Trauma Score
(RTS), Injury Severity Score (ISS), on call, time from arrival at the
hospital until first computed tomography (CT), days in the ICU, ASA
preinjury and survival after 30 days.

@@

OFI was identified in 143 of 1449 patients. Revised Trauma Score, ≤ 7
days in the ICU and death within 30 days after hospitalization were
significantly associated with OFI in the unadjusted analysis, and
Revised Trauma Score and ≤ 7 days in the ICU were significantly
associated with OFI in the adjusted analysis.

@@

The patient and process factors associated with OFI indicated that
higher RTS was associated with higher odds of OFI, and ≤ 7 days in the
ICU was associated with lower odds of OFI.

@@

RR = respiratory rate

@@

SBP = systolic blood pressure

@@

RTS is a physiological scoring system scored from the data obtained in
the first set of data obtained on the patient and consists of Glasgow
Coma Scale (GCS), respiratory rate and systolic blood pressure. RTS is
heavily weighted towards GCS compared to respiratory rate and systolic
blood pressure. It has been demonstrated to show accuracy at predicting
death. Each of the mentioned variables get a corresponding code value
between 0 and 4 based on the value of GCS, RR and SBP. The formula for
RTS is RTS = (0.9368 x GCS code value) + (0.7326 x SBP code
value) + (0.2908 x RR code value) [@20].

@@

We conducted a retrospective single-centre cohort study based on data
from the Karolinska University Hospital in Solna and from the Swedish
Trauma registry. We used R for all statistical analyses [@19].

For the RTS-value the GCS, SBP and RR values were converted to a
RTS-code between 0 and 4.

For patients where hospital values for systolic blood pressure,
respiratory rate or GCS score were missing, prehospital values were used
instead. If the patient was intubated A complete case analysis was
conducted after handling missing values. We present sample
characteristics using descriptive statistics.

@@

The quantitative variables were factors chosen from the trauma registry,
based on the locally used audit filters, standard epidemiological
factors and factors registered in the Swedish Trauma Registry. The
categorical factors were sex, age, if the patient was intubated or not
and how many days, Revised Trauma Score (RTS), Injury Severity Score
(ISS), on call, time from arrival at the hospital until first computed
tomography (CT), days in the ICU, ASA preinjury and survival after 30
days. On duty was defined as before 8.00 a.m., after 5 p.m., or during a
weekend - defined as Saturday or Sunday.

@@

@@

# Results

```{r, include=FALSE}
source("functions/flowchart.R")
```

```{r, echo=FALSE}
for (i in 1:(length(vert) - 1)) {
  connectGrob(vert[[i]], vert[[i + 1]], type = "vert") %>%
    print
}

@@

## Descriptive data

```{r include_scripts, include=FALSE}
source("functions/table1.R")
```

```{r, echo=FALSE}
table2
```

## Main results

```{r, include=FALSE}
source("functions/table2.R")
```

```{r, echo=FALSE}
table3_merge
```

14,022 patients were included in the trauma registry and trauma care
quality database between 2017 and 2023. 12,278 patients were excluded,
leaving a total of 1,742 patients eligible for the study. 11,343 were
excluded because they were not treated in the ICU. 3 patients were
excluded because they were under 15 years old. 6 patients were excluded
because they were dead on arrival, and 928 patients were excluded
because there was no data on the presence or absence OFI. 293 patients
were excluded because there was missing data, leaving a total of 1,449
patients included in the study.

Table 1 present sample characteristics. 1,096 (76%) patients were male
and 353 (24%) patients were female. 846 (58%) patients had no
intubation, 241 (17%) patients were intubated for 1-7 days and 113
(7,8%) patients were intubated for more than 7 days. The number of
patients who arrived outside working hours --- on Saturday or Sunday or
after 5 pm or before 8 am were 1,042 (72%). 1,064 (73%) patients were
equal to or less than 7 days in the ICU, and 385 (27%) patients were
more than 7 days in the ICU. 1,080 (75%) patients had an ASA-score of
1-2 before arrival to the hospital, and 369 (25%) patients had an
ASA-score of 3-7 before arrival to the hospital. 286 (20%) patients died
within 30 days after hospitalization.

@@

RTS was significantly associated with increased odds of OFI in both the
unadjusted and adjusted analyses. ≤ 7 days in the ICU was significantly
associated with decreasing odds of OFI in both the unadjusted and
adjusted analyses.

Sex, Age, Intubation, ISS, time to first CT, on duty and ASA preinjury
were not significantly associated with OFI in the unadjusted or the
adjusted analyses. Female sex, intubation \> 7 days, \> 7 days in the
ICU, ASA preinjury 1-2 and survival 30 days after hospitalisation could
not be analysed with logistic regression.

@@

We found that higher RTS was associated with higher odds of OFI. RTS is
a combined value of systolic blood pressure, Glasgow Coma Scale and
respiratory rate with more weight towards GCS as mentioned in the
introduction. An explanation for the increased odds of OFI in patients
with increased RTS could be that the patients with low RTS values - with
low GCS, RR and SBP also are the once that are in a bad state upon
arrival - upon which emergency procedures e.g. intubation have been
carried out rapidly due to the patients being very ill from the
beginning. The patients who are being missed and where certain
procedures take longer time before being performed are the patients who
seem stable upon arrival, but who later become worse --- the patients
with higher RTS scores. In Albaaj's study respiratory rate \>29,
systolic blood pressure 50-75 mmHg and 76-89 mmHg, and Glasgow Coma
Scale 9-12 were significantly related to increased odds of OFI in the
unadjusted analysis, but not in the adjusted analysis [@12]. All of
these variables equal medium RTS values.

The variable ≤ 7 days in the ICU was significantly associated with
decreasing odds of OFI in both the adjusted and unadjusted analyses. An
explanation would be that less days spent in the ICU compared to more
days spent in the ICU indicates that patients have been in a better
state and needed less time of recovery before leaving the hospital.
Furthermore, a longer stay leads to more space for mistakes being made.

Death within 30 days after hospitalization was associated with OFI in
the unadjusted analysis but not in the adjusted analysis. Patients who
died within 30 days after hospitalization could either have suffered a
severe trauma or have been in a poor state from the beginning; therefore
increasing the odds of OFI.

@@

There were several limitations in our study: the selection for the
morbidity and mortality review, the definition of the RTS variable and
the study size.

The selection for review for the morbidity and mortality review relied
mostly on the local audit filters used at the Karolinska Institute
Hospital in Solna. This leads to a risk of misplacement where some
patients with OFI were not selected by for the review based on the audit
filters although there was an OFI later on in the hospitalization
process. #Lägga in källa för audit filters - Kelvins artikel?

The definition of the RTS-value is a combination of GCS, SBP and RR.
When the patients were intubated and had missing values for GCS or RR,
we replaced it with RTS with a value of 0 for the equation for RTS, as
GCS and respiratory rate are 0 when patients are intubated. This could
affect the final results, because then we would not get their initial
values for these variables, meaning some patients received a RTS score
more weighted towards SBP than they should have. However, the use of RTS
can also be seen as an advantage, being a combination of vital
parameters. This would mean that the RTS value reflects the state of the
patient better than these parameters separately.

The study was a single-centre study and illustrate the situation in the
Karolinska Intitute Hospital in Solna. More research is needed on the
area of OFI in patients admitted to the ICU. A suggestion would be to
conduct a multi-centre study on this group of patients to further
investigate areas for improvement and to further develop trauma quality
programs. Based on the results from this study, variables of interest
could be number of days in the ICU and investigation of GCS, RR and SBP
separately instead of them as a combined RTS-score. A study including
more patients could potentially show significance for the separate vital
parameters leading to a better prediction model for which patients that
should receive more focus when improving trauma care.

@@

The study is conducted on 143 outcomes of OFI. There were few outcomes
because we included only those who were admitted to the ICU, \> 14 years
old, not dead on arrival and had a registration for the presence or
absence of OFI. The external validity is therefore compromised due to
the limited size of the Swedish database.

Based on the statistical significance for higher RTS being associated
with higher odds of OFI, there should be more focus on trauma patients
who are stable upon arrival who have a risk of becoming worse. Future
studies should investigate what factors to use to see the indication for
such a focus. Focusing on patients who are stable upon arrival overall
is very extensive. Therefore, identifying such factors would be
valuable.

@@

The aim of this study was to characterize patient and process factors
among adult patients admitted to the ICU that were associated with OFI.
The study showed that higher RTS score is associated with increased odds
of OFI indicating that patients who are relatively well upon arrival to
the hospital are the group where improved trauma care should be focused
on. The logistical regression also showed that ≤ 7 days in the ICU was
associated with decreased odds of OFI indicating that patients who are
at the hospital less days in the ICU compared to more days in the ICU
are less prone to opportunities for improvement.
